Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-05-07 writing/manuscript.Rmd
added tucker 2018
addressed most of Mike's comments

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels. Previous research has quantified the effects of temperature on mammalian movement behavior, but few have quantified the effects of future climate change on fine-scale movement and habitat selection. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammals' movement rates and habitat selection. We then leverage the estimated effects to predict how and where mammals will move in British Columbia (BC), Canada, throughout the 21^st^ century. We show that changes in temperature altered movement phenology, and that climate change will impact mammals' movement rates and habitat selection throughout the century, but that there are no common trends across all study species. Consequently, we suggest that species' responses be interpreted carefully and considered jointly when extrapolating the community-level consequences of climate change."
```

@@

@@

<!-- place table of contents on a new page -->
\newpage

\tableofcontents

@@

\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Additionally, temperature also affects movement indirectly by through other drivers of movement behavior, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023]. But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024]. However, generalists' grater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to adapt within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021]. Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].

@@

Uncertainty in future conditions and their consequences on ecosystems presents a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behavior and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024]. In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare for and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [e.g., @tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [e.g., @zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future. Failing to understand how changes in temperature will affect mammalian movement behavior and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data thoughout a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behavior of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.

@@

```{r data-table}
dt_data <-
  readRDS('../models/movement-models-2024-06-06.rds') %>%
  transmute(
    Species = case_when(
      dataset_name == 'Canis_lupus_boreal' ~ '\\emph{Canis lupus}',
      dataset_name == 'Rangifer_tarandus_boreal' ~ '\\emph{Rangifer tarandus} (boreal)',
      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (s. mountain)',
      dataset_name == 'Puma_concolor_2' ~ '\\emph{Puma concolor}',
      dataset_name == 'Puma_concolor_4' ~ '\\emph{Puma concolor}',
      dataset_name == 'Elk in southwestern Alberta' ~ '\\emph{Cervus canadensis}',
      dataset_name == 'Oreamnos_americanus' ~ '\\emph{Oreamnos americanus}',
      dataset_name == 'Ursus_arctos_horribilis' ~ '\\emph{Ursus arctos horribilis}'),
    animal, # keep animal column for grouping later
    tel = map(tel, data.frame),
    na_speed = map_lgl(movement_model,
                       \(.m) suppressWarnings(speed(.m)$DOF == 0))) %>%
  unnest(tel) %>%
  group_by(Species, animal) %>%
  summarize(Start = min(as.Date(timestamp)),
            End = max(as.Date(timestamp)),
            dt = median(timestamp - lag(timestamp), na.rm = TRUE),
            animal = unique(animal),
            na_speed = unique(na_speed),
            .groups = 'drop') %>%
  group_by(Species) %>%
  summarize(Start = min(Start),
            End = max(End),
            dt = as.numeric(median(dt, na.rm = TRUE)),
            Animals = n_distinct(animal),
            n_missing = sum(na_speed),
            .groups = 'drop') %>%
  mutate(dt = round('hours' %#% dt, 2)) %>% # convert from seconds to hours
  rename('\U0394t (hours)' = dt,
         'NA speed' = n_missing)

knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
```

@@

This manuscript leverages four main datasets: (1) a collection of GPS telemetry data; (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. We detail the data sources and analyses in the sections below.<!--, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).-->

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior. The script used to clean the data and all associated custom functions are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

We simulated hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. We then predicted changes in movement behavior and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile<!-- (Appendix C)-->.

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).

Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig. S3).

@@

@@

\noindent In this manuscript, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behavior by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behavior (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).

@@

\noindent We estimated the effects of temperature on mammals' movement state (moving or not) and speed (when moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017]. The first HGAM estimated the probability that an animal was moving with a binomial family of distributions and logit link function. The second HGAM estimated an animal's speed (when moving) with a gamma family of distributions and log link function. The HGAMs included random intercepts for each animal (`s(..., bs = 're')`), fixed-effect intercepts for each species, and species-level `by` smooths for time of day, day of year, and temperature [`s(..., by = species)`, see model I in Figure 4 of @pedersen_hierarchical_2019]. Additionally, the models had three tensor product interaction terms (`ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year. These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing). The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, @wood_generalized_2017]. Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving. We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020]. Additional details are provided in Appendix A.

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from water by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B, also see @aarts_estimating_2008]. We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [*sensu* @johnson_comparison_1980]. While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, prey availability), we decided to only use these three proxies to produce results that are comparable between species and to make province-wide predictions simpler. Each species' model had the same structure:

@@

```{r rsf, eval=FALSE, echo=TRUE}
rsf <- bam(
  detected ~ # 1 for telemetry locations, 0 for quadrature data
    # species-level average resource preference
    s(forest_perc, k = 6, bs = 'tp') +
    s(elevation_m, k = elev_k, bs = 'tp') +
    s(dist_water_m, k = 6, bs = 'tp') +
    # animal-level deviations from the species-level average
    s(animal, bs = 're') +
    s(forest_perc, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
    s(elevation_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
    s(dist_water_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
    # changes in preference with temperature
    ti(forest_perc, temperature_C, k = 6, bs = 'tp') +
    ti(elevation_m, temperature_C, k = 6, bs = 'tp') +
    ti(dist_water_m, temperature_C, k = 6, bs = 'tp') +
    # include marginals of temperature to remove sampling biases
    s(temperature_C, k = 4, bs = 'tp') +
    s(temperature_C, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
  family = poisson(link = 'log'),
  data = d,
  weights = weight, # based on AKDE
  method = 'fREML',
  discrete = TRUE)
```

@@

At average temperatures, wolves preferred dense forest cover ($\gtrapprox$ 50%) high elevations, and distances from water < 10 km; elk preferred intermediate forest cover, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats preferred sparse (< 50%) forest cover, elevations between 1 and 2 km; pumas preferred high forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to high forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.

>>>>

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels. Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on fine-scale movement and habitat selection. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammals' movement rates and habitat selection. We then leverage the estimated effects to predict how and where mammals will move in British Columbia (BC), Canada, throughout the 21^st^ century. We show that changes in temperature altered movement phenology, and that climate change will impact mammals' movement rates and habitat selection throughout the century, but that there are no common trends across all study species. Consequently, we suggest that species' responses be interpreted carefully and considered jointly when extrapolating the community-level consequences of climate change."
```

@@

<!-- using  "behaviOR" in key words to increase audience -->

@@

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Additionally, temperature also affects movement indirectly by through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023]. But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while thermal generalists can adapt more easily to change [@dupont_beyond_2024]. However, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to respond within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@ipcc_climate_2023; @zurowski_summer_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. Consequently, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare for and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [e.g., @tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [e.g., @zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data thoughout a large spatial range of British Columbia and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.

@@

```{r data-table}
dt_data <-
  readRDS('../models/movement-models-2024-06-06.rds') %>%
  transmute(
    Species = case_when(
      dataset_name == 'Canis_lupus_boreal' ~ '\\emph{Canis lupus}',
      dataset_name == 'Rangifer_tarandus_boreal' ~ '\\emph{Rangifer tarandus} (boreal)',
      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (s. mountain)',
      dataset_name == 'Puma_concolor_2' ~ '\\emph{Puma concolor}',
      dataset_name == 'Puma_concolor_4' ~ '\\emph{Puma concolor}',
      dataset_name == 'Elk in southwestern Alberta' ~ '\\emph{Cervus canadensis}',
      dataset_name == 'Oreamnos_americanus' ~ '\\emph{Oreamnos americanus}',
      dataset_name == 'Ursus_arctos_horribilis' ~ '\\emph{Ursus arctos horribilis}'),
    animal, # keep animal column for grouping later
    tel = map(tel, data.frame),
    na_speed = map_lgl(movement_model,
                       \(.m) suppressWarnings(speed(.m)$DOF == 0))) %>%
  unnest(tel) %>%
  group_by(Species, animal) %>%
  summarize(Start = min(as.Date(timestamp)),
            End = max(as.Date(timestamp)),
            dt = median(timestamp - lag(timestamp), na.rm = TRUE),
            animal = unique(animal),
            na_speed = unique(na_speed),
            .groups = 'drop') %>%
  group_by(Species) %>%
  summarize(Start = min(Start),
            End = max(End),
            dt = as.numeric(median(dt, na.rm = TRUE)),
            Animals = n_distinct(animal),
            n_missing = sum(! na_speed),
            .groups = 'drop') %>%
  mutate(dt = round('hours' %#% dt, 2)) %>% # convert from seconds to hours
  rename('\U0394t (hours)' = dt,
         'Has speed' = n_missing)

knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with resolved speed (Has speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
```

@@

This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig. \@ref(fig:bc-map)); (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. We detail the data sources and analyses in the sections below.<!--, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).-->

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA*** Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour. The script used to clean the data and all associated custom functions are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

We simulated hourly variation in future weather by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. We then predicted changes in movement behaviour and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile<!-- (Appendix C)-->.

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 animals with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).

Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig. S3).

@@

keeping for thesis:

@@

\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).
-->

@@

\noindent We estimated the effects of temperature ($^\circ$C) on mammals' movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017]. The first HGAM estimated the probability that an animal was moving, $P(M)$, with a binomial family of distributions and logit link function. The response variable `moving` was coded as 1 if moving and 0 if not. The second HGAM estimated an animal's speed when moving (`speed`) with a gamma family of distributions and log link function. The HGAMs included random intercepts for each animal (`s(animal, bs = 're')`), fixed-effect intercepts for each species (`species`), and species-level `by` smooths (`s(..., by = species)`), which allowed each species' term to be estimated independent of other species [see model I in Figure 4 of @pedersen_hierarchical_2019]. The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`). Additionally, the models had three tensor product interaction terms (`ti()`) `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year. These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing). Finally, smooth terms of log-transformed hours between GPS locations (`dt`) accounted for irregular sampling rates. A global term of `log(dt)` accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`s(log(dt), species, bs = 'fs')`) allowed for species-level deviations from the global term while assuming a common smoothness parameter across species [see model GS in Figure 4 of @pedersen_hierarchical_2019]. The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`$;$ see p. 202 of @wood_generalized_2017]. Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving. We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020]. Additional details are provided in Appendix A.

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B, also see @aarts_estimating_2008]. We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [*sensu* @johnson_comparison_1980]. While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, prey availability), we decided to only use these three proxies to produce results that are comparable between species and to make province-wide predictions simpler. Each species' model had the same structure:

@@

```{r rsf, eval=FALSE, echo=TRUE}
rsf <- bam(
  detected ~ # 1 for telemetry locations, 0 for quadrature points
    # species-level average resource preference
    s(forest_perc, k = 6, bs = 'tp') +
    s(elevation_m, k = elev_k, bs = 'tp') +
    s(dist_water_m, k = 6, bs = 'tp') +
    # animal-level deviations from the species-level average
    s(animal, bs = 're') +
    s(forest_perc, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
    s(elevation_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
    s(dist_water_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
    # changes in preference with temperature
    ti(forest_perc, temp_c, k = 6, bs = 'tp') +
    ti(elevation_m, temp_c, k = 6, bs = 'tp') +
    ti(dist_water_m, temp_c, k = 6, bs = 'tp') +
    # include marginals of temperature to remove sampling biases
    s(temp_c, k = 4, bs = 'tp') +
    s(temp_c, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
  family = poisson(link = 'log'),
  data = d, # species-specific dataset
  weights = weight, # based on AKDE
  method = 'fREML',
  discrete = TRUE)
```

@@

Near 0$^\circ$C, wolves preferred dense forest cover ($\gtrapprox$ 50%) high elevations, and distances from water < 10 km; elk preferred intermediate forest cover, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats preferred sparse (< 50%) forest cover, elevations between 1 and 2 km; pumas preferred high forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to high forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.
